---
title: "R Basics and Data Management"
author: "Your Name"
date: "2025-01-21"
output: html_document
---

## Introduction

This R Markdown document demonstrates best practices, basic concepts, and data management techniques in R using a school enrollment dataset.

------------------------------------------------------------------------

## 1. Setting Up the Environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)  # For reproducibility
```

------------------------------------------------------------------------

```{r}

```

```{r}
write.csv(student_data="student_dummy.csv")
```

## 2. Creating the Dataset

```{r create-dataset}
# Generate the dataset
n <- 290  # Number of observations
library(dplyr)

student_data <- data.frame(
  Student_ID = sprintf("S%03d", 1:n),
  Name = paste0("Student_", 1:n),
  Age = sample(13:17, n, replace = TRUE),
  Gender = sample(c(1, 2), n, replace = TRUE),
  Grade_Level = sample(8:12, n, replace = TRUE),
  Enrollment_Type = sample(c("Full-time", "Part-time"), n, replace = TRUE, prob = c(0.7, 0.3)),
  Subjects_Enrolled = sample(3:7, n, replace = TRUE),
  Attendance_Rate = round(runif(n, 70, 100), 1),
  Average_Score = round(rnorm(n, mean = 80, sd = 10), 1)
)

# Introduce some missing values
missing_indices <- sample(1:n, size = 20)
student_data$Attendance_Rate[missing_indices] <- NA

```

------------------------------------------------------------------------

## 3. Renaming Columns

```{r renaming-columns}
# Renaming columns
student_data <- student_data %>%
  rename(
    ID = Student_ID,
    Score = Average_Score
  )
head(student_data)
```

------------------------------------------------------------------------

## 4. Recoding Values

```{r recoding-values}
# Recode Gender and Grade_Level
student_data <- student_data %>%
  mutate(
    Gender = recode(Gender, "Male" = 1, "Female" = 2),
    Grade_Label = recode(Grade_Level,
                         `8` = "Middle School",
                         `9` = "Freshman",
                         `10` = "Sophomore",
                         `11` = "Junior",
                         `12` = "Senior")
  )
head(student_data)
```

------------------------------------------------------------------------

## 5. Handling Missing Data

### Dropping Missing Values

```{r drop-na}
# Dropping rows with missing values
cleaned_data <- student_data %>% drop_na()
dim(cleaned_data)
```

### Replacing Missing Values

```{r replace-na}
# Replace missing Attendance_Rate with mean
student_data <- student_data %>%
  mutate(Attendance_Rate = ifelse(is.na(Attendance_Rate), mean(Attendance_Rate, na.rm = TRUE), Attendance_Rate))
head(student_data)
```

------------------------------------------------------------------------

## 6. Filtering and Summarizing Data

### Filtering Data

```{r filtering}
# Filter Full-time students
full_time_students <- student_data %>%
  filter(Enrollment_Type == "Full-time")
head(full_time_students)
```

### Summarizing Data

```{r summarizing}
# Summary statistics
summary_stats <- student_data %>%
  group_by(Grade_Level, Gender) %>%
  summarize(
    Mean_Score = mean(Score, na.rm = TRUE),
    Median_Attendance = median(Attendance_Rate, na.rm = TRUE),
    Total_Students = n()
  )
summary_stats
```

------------------------------------------------------------------------

## 7. Data Visualization

```{r visualization}
library(ggplot2)
# Boxplot of Scores by Grade Level
ggplot(student_data, aes(x = factor(Grade_Level), y = Score, fill = as.factor(Gender))) +
  geom_boxplot() +
  labs(
    title = "Average Scores by Grade Level and Gender",
    x = "Grade Level",
    y = "Average Score"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 8. Exporting Cleaned Data

```{r export-data}
# Export the cleaned data to a CSV
write.csv(student_data, "cleaned_student_data.csv", row.names = FALSE)
```

------------------------------------------------------------------------

## Conclusion

This document covered R basics, data cleaning, and data management techniques. The steps included renaming columns, handling missing values, recoding variables, and visualizing data. The cleaned dataset has been saved for further analysis.
